### 2017-09-13 chapters/introduction/mechanics.tex
Update to latest version

\begin{align*}
  \int_{\partial \Omega} \mathbf{t} \mathrm{d}s
  = \int_{\partial \Omega} \sigma \mathbf{n} \mathrm{d}s
  = \int_{\Omega} \nabla \cdot \sigma \mathrm{d}v,
\end{align*}
and by collecting the terms from \eqref{eq:cons_lin_mom}. We arrive at
Cauchys momentum equation
\begin{align}
  \nabla \cdot \sigma + \mathbf{b} =  \rho \dot{\mathbf{v}}.
  \label{eq:chauch_momentum_eq}
\end{align}
The contribution from the body force ($\mathbf{b}$)  and inertial term
($\rho \dot{\mathbf{v}}$) are negligible compared to the stresses
\cite{hunter1996kd,tallarida1970left, moskowitz1981effects}, which is
why the force balance equations is typically only stated as
\begin{align}
  \nabla \cdot \sigma = \mathbf{0}.
  \label{eq:momentum_simple_current}
\end{align}
Note that we have formulated the balance law in the current
configuration. An equivalent statement can be formulated in terms of
the reference configuration
\begin{align}
  \nabla \cdot \FPK = \mathbf{0}, 
  \label{eq:momentum_simple_reference}
\end{align}
where $\FPK$ is the first Piola-Kirchhoff stress tensor.

@@

\subsubsection{General requirements for the strain-energy density
  function}
\label{sec:strain_energy_req}
Some general requirements must hold for the strain-energy function.
First of all, we require that the reference state is stress free and
that the stored energy increases monotonically with the deformation. 
Formally this can be stated simply as 
\begin{align*}
  \Psi(\I) = 0 \; && \text{and} &&\; \Psi(\F) \geq 0.
\end{align*}
Moreover, expanding or compressing a body to zero volume would
require an infinite amount of energy, i.e
\begin{align*}
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& 0 \\
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& \infty
\end{align*}
We say that the strain energy should be objective, meaning that the
stored energy in the material should be invariant with respect to
change of observer. Formally we must have: \emph{given any positive symmetric
rank-2 tensor $\C \in \mathrm{Sym}$:}
\begin{align}
  \Psi(\mathbf{C}) = \Psi(\mathbf{Q}\mathbf{C}\mathbf{Q}^T), \; \forall \mathbf{Q} \in \mathcal{G} \subseteq \mathrm{Orth}.
\end{align}
Here $\mathrm{Orth}$ is the group of all positive orthogonal matrices.
If $\mathcal{G} = \mathrm{Orth}$ we say that the material is
isotropic, and otherwise we say that the material is anisotropic.
This brings us to another important issue, which is related to the
choice of coordinate-system. Having to deal with different
coordinate-systems, and mapping quantities from one coordinate-system
to another can results in complicated computations. Therefore it would be beneficial if we
could work with quantities which do not depend on the choice of
coordinate-system. Such quantities are called invariants. 
If the material is isotropic, the representation theorem for
invariants states that $\Psi$ can be expressed in terms of the
principle invariants of $\mathbf{C}$, that is $\Psi = \Psi(I_1, I_2,
I_3)$. The principle invariants $I_i, i=1,2,3$ are the coefficients in
the characteristic polynomial of $\mathbf{C}$, and is given by 
\begin{align}
  I_1 = \tr \C,  && I_2 = \frac{1}{2}\left[ I_1^2 - \tr(\C^2)\right] && \text{and} && I_3 = \det \C.
\end{align}
In the case when the material constitutes a transversely isotropic
behavior, that is the material has a preferred direction $\mathbf{a}_0$,
which in the case of the myocardium could be the direction of fiber
muscle fibers, we have
\begin{align*}
  \mathcal{G} = \{ \mathbf{Q} \in \mathrm{Orth}: \mathbf{Q}(\mathbf{a}_0\otimes\mathbf{a}_0)\mathbf{Q}^T\},
\end{align*}
with $\otimes$ being the outer product. In this case the strain-energy
density function can still be expressed through invariants. However,
we need to include the so called quasi-invariants, which are defined
as stretches in the local microstructural coordinate-system. The
transversely isotropic invariants are given by
\begin{align*}
  I_{4\mathbf{a}_0 } = \mathbf{a}_0 \cdot (\C \mathbf{a}_0) && \text{and} && I_5 = \mathbf{a}_0 \cdot (\C^2 \mathbf{a}_0).
\end{align*}
The invariants do have a physical interpretation. For instance, $I_3$
is related to the volume ratio of material during deformation, while
$I_{4\mathbf{a}_0 } $ is related to the stretch along the direction
$\mathbf{a}_0 $. Indeed the \emph{stretch} ratio in the direction
$\mathbf{a}_0$ is given by $\lambda_{\mathbf{a}_0} = | \F \mathbf{a}_0
|$ and we see that $I_{4\mathbf{a}_0 }  =  \mathbf{a}_0 \cdot (\C
\mathbf{a}_0) = \F \mathbf{a}_0 \cdot (\F \mathbf{a}_0) =
\lambda_{\mathbf{a}_0}^2$. For more details about invariants see
\cite{holzapfel2009constitutive,liu1982representations}.

The theory of global existence of unique solutions for elastic problem
was originally based convexity of the free energy function.
A function $\phi: \Omega \rightarrow \mathbb{R}$ is \emph{convex} if for any
$\Xvec_1, \Xvec_2 \in \Omega$ we have
\begin{align}
  \phi(\lambda \Xvec_1+ (1-\lambda) \Xvec_2)
  \leq \lambda \phi(\Xvec_1)
  + (1-\lambda) \phi(\Xvec_2).
\end{align}
However, from a physical point of view this requirement is too strict
\cite{ball1976convexity}. A slightly weaker requirement is that the
strain-energy function $\Psi$ is \emph{polyconvex}, meaning that there exist
a convex function $\phi$ such that
\begin{align*}
  \Psi(\F) = \phi(\F, \cof \F, \det \F).
\end{align*}
Note however, that if $\Psi$ is convex then it is also polyconvex.

@@

Another approach which is used in this thesis is to impose a Robin
type boundary condition at the base
\begin{align}
  \FPK \Nvec + k \uvec = \mathbf{0},  \;  \Xvec \in \lvbase, 
\end{align}
or at the epicardium to mimic the pericardium
\begin{align}
  \FPK \Nvec + k \uvec = \mathbf{0},  \;  \Xvec \in \lvepi, 
\end{align}
Here $k$ can be seen as the stiffness of a spring that limit the
movement. The limiting cases, $k = 0$ and $k \rightarrow
\infty$ represent free and fixed boundary respectively.
More complex boundary conditions to mimic the pericardium is also
possible \cite{fritz2014simulation}, but not considered in this thesis.
An overview of the location of the different boundaries for the
bi-ventricular geometry is illustrated in Figure \ref{fig:boundaries}.

@@

\subsection{Force-balance equation}
We will now collect all the terms that makes up to force balance for
the cardiac mechanics problem. Considering the myocardium as a incompressible,
hyperelastic material with a Robin type boundary condition at the base
we obtain the following strong form in the Lagrangian formulation
\begin{align}
  \begin{split}
  \nabla \cdot \FPK &= 0 \\
  J - 1 &= 0,
  \end{split}
 \label{eq:force_balance_strong}
\end{align}
completed with appropriate boundary conditions. To solve this
numerically using the finite element method, we need to derrive weak
variational form of this equaiton.
% \begin{align}
%   \FPK \Nvec + k \uvec &= \mathbf{0},  \;  \Xvec \in \lvbase\\
%   u_1 &= 0,  \;  \Xvec \in \lvbase \\
%   \FPK\mathbf{N}  &= \mathbf{0}, \;  \Xvec \in \epi \\
%   \FPK\mathbf{N} &= -p_{\mathrm{lv}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \lvendo
%   \label{eq:bndry_conditions_strong}
% \end{align}

@@

external energy. Internal energy, is energy that is stored within the

@@

For a hyperelastic and incompressible material, the total potential

@@

Equation \eqref{eq:force_balance_strong} is only possible to solve
analytically for very simplified problems. Therefore we need to employ
a numerical approximation method to solve the problem. One such method
is the finite element method (FEM). When using the finite element method, we often refer to such
approximation as a Galerkin approximation. This is based on
approximating the solution by linear combinations of  basis functions in a finite dimensional
subspace of the true solution. If $V$ and $Q$ are two suitable Hilbert
spaces for the displacement $\uvec$ and the hydrostatic pressure $p$
respectively (see Section \ref{sec:variational_formulation}), we now
select some finite dimensional subspaces $V_h \subset V$ and $Q_h
\subset Q$, which are spanned by a finite number of basis functions.

@@

\begin{remark}
  The basis functions that span the Taylor-Hood
  finite element spaces are also known as the
  Lagrangian basis functions. These basis functions, of
  degree $n$, are a polynomials of degree $n$ on each element, but only
  continuous at the nodes (i.e not continuously
  differentiable). Consequently, differentiating a function that is
  expressed as a linear combination of the Lagrangian basis functions,
  will not be continuous at the nodes, and therefore caution has to
  be made when evaluating features that depends on the derivative of
  such functions. Examples of such features are stress and
  strain with depends on the deformation gradient which again depends
  on the derivative of the displacement. One way to deal with this
  issue is to 1) use other types of elements that are continuously
  differentiable everywhere,  such a the cubic Hermite elements or 2) evaluate the features at the
  Gaussian quadrature points where there is no problem with continuity.
\end{remark}

\subsection{Constitutive relations}
\label{sec:constitutive_relations}
We have now covered a mechanical framework which holds any for
material in general. What differentiate the mechanics of soft
living tissue, like the myocardium, from other materials is the
constitutive relations which describes the response of a material to
applied load. Such constitutive relations often comes from
experimental observations, both on observation of anatomical structure
but also from experiments done on tissue slabs.

@@

In this thesis we have in several occasions used a
transversely isotropic version of \eqref{eq:holzapel_full} which is
obtained by setting $a_{fs} = b_{fs}= a_s = b_s = 0$, i.e
\begin{align}
  \label{eq:holzapel_trans}
  \begin{split}
  \Psi(I_1, I_{4\ef}) = \frac{a}{2 b} \left( e^{ b (I_1 - 3)}  -1 \right)
  + \frac{a_f}{2 b_f} \left( e^{ b_f (I_{4\ef} - 1)_+^2} -1 \right).
  \end{split}
\end{align}
If we further set $a_f = b_f = b = 0$ so that in $a$ is the
only nonzero parameter, then the Holzapfel-Ogden model reduces (after a
series expansion of the exponential and a limiting argument) to
\begin{align}
  \Psi(I_1)  = \frac{a}{2} \left( I_1 - 3 \right), 
\end{align}
which is the model of a Neo Hookean material. The Cauchy stress can be derived
analytically from \eqref{eq:holzapel_full}, by using the chain rule and
\eqref{eq:cauchy_incomp}, 
\begin{align}
  \begin{split}
    J\Cauchy
    =& \frac{\partial \Psi(\F)}{\partial \F}\F^{T} + p \I
    = \sum_{i \in \left\{ 1, 4\ef,  4\eS,  8\ef\eS \right\} }
    \psi_i \frac{\partial I_i}{\partial \F}\F^{T} + p \I \\
    =& p \I + a \left( e^{ b (I_1 - 3)}  -1 \right) \mathbf{B} 
    + 2 a_f (I_{4\ef} - 1)_+  e^{ b_f (I_{4\ef} - 1)^2} \mathbf{f} \otimes \mathbf{f} \\
    &+ 2 a_f (I_{4\eS} - 1)_+  e^{ b_f (I_{4\eS} - 1)^2} \mathbf{s} \otimes \mathbf{s} 
    + a_{fs} I_{8\ef\eS}  e^{ b_{fs} I_{8\ef\eS}^2} \left( \mathbf{f} \otimes \mathbf{s}  +  \mathbf{s} \otimes \mathbf{f} \right),
  \end{split}
\end{align}
where $\mathbf{B} = \F \F^{T}$ is the left Cauchy-Green tensor,
$\mathbf{f} = \F \ef$ and $\mathbf{s} = \F \eS$.

@@

The active stress approach is based one the classical three element
Hill model illustrated in Figure \ref{fig:hill_muscle_model}, where the
active contribution naturally decomposes the total stress into a sum
of passive and active stresses
\cite{nash2004electromechanical}. Hence, in the active stress
formulation \cite{hunter1998modelling} one assumes that the total
Cauchy stress $\Cauchy$ can be written as an additive sum of one
passive contribution $\Cauchy_p$ and one active contribution $\Cauchy_a$,
\begin{align}
  \Cauchy = \Cauchy_p + \Cauchy_a && J \Cauchy_p =  \frac{\partial \Psi(\F)}{\partial \F} \F^{T}.
\end{align}
The passive contribution is determined by the material model used,
while the active contribution is given by 
\begin{align}
  \Cauchy_a = \Cauchy_{ff} \Fef \otimes \Fef +
  \Cauchy_{ss} \mathbf{s} \otimes \mathbf{s} +
  \Cauchy_{nn} \mathbf{n} \otimes \mathbf{n},
\end{align}
and the different constants $\Cauchy_{ff}, \Cauchy_{ss}$, and
$\Cauchy_{nn}$, which are the active stress in the fiber, sheet and
sheet-normal direction respectively, are typically coupled to the
electrophysiology and calcium dynamics.
There are experimental evidence that the active stresses in the
transverse direction of the fibers ($\Cauchy_{ss}$, and $\Cauchy_{nn}$),
are non-negligible \cite{lin1998multiaxial}, and one approach is to assume
a uniform transverse activation in which the total active tension
can be written as 
\begin{align}
  \Cauchy_a = T_a \left[\Fef \otimes \Fef +
   \eta\left( \mathbf{s} \otimes \mathbf{s} +
  \ \mathbf{n} \otimes \mathbf{n} \right)\right],
  \label{eq:intro_active_stress}
\end{align}
where $\eta$ represent the amount of transverse activation and $T_a
\in \mathbb{R}$ is the magnitude of the active tension.
In the limiting case ($\eta = 0.0$), the active tension acts purely
along the fibers and is reduced to 
\begin{align}
  \Cauchy_a = T_a \Fef \otimes \Fef.
\end{align}
Note that, by observing  that
\begin{align*}
  \frac{\partial I_{4\mathbf{a}_0}}{\partial \F} = \frac{\partial (\mathbf{a}_0  \cdot \C \mathbf{a}_0 )}{\partial \F}
  = 2 \mathbf{a} \otimes \mathbf{a}_0 \implies  \mathbf{a} \otimes  \mathbf{a}= \frac{1}{2} \frac{\partial I_{4\mathbf{a}_0}}{\partial \F} \F^{T}
\end{align*}
and that $I_1 =  I_{4\ef} +  I_{4\eS} +  I_{4\en}$, 
we can instead decompose the strain-energy function into a passive and active
parts \cite{pathmanathan2010cardiac}, $\Psi= \Psi_p + \Psi_a$, with
\begin{align}
\Psi_a = \frac{T_a}{2J} \left(( I_{4\ef} - 1)  + \eta \left[ (I_1 - 3) -
    (I_{4\ef} - 1)\right] \right), 
\end{align}
so that $J \Cauchy_a  = \frac{\partial \Psi_a}{\partial \F}
\F^{T}$.

@@

The active part $\F_a$, is an inelastic process driven by the
biochemistry and can be seen as the actual distortion of the
microstructure. The elastic part $\F_e$ is responsible for preserving
compatibility of the tissue and stores all the energy in the
deformations. As a consequence, the strain energy function is a
function of the elastic deformation gradient only $\F_e$. The
decoupling can illustrated by considering two sarcomeres connected in
series as shown in Figure \ref{fig:actstrain}.

@@

The FEniCS project is an open-source computing platform for solving
partial differential equations (PDEs) using the finite element method (FEM).
Solving PDEs using FEM involves many implementation details that can
be tedious to implement yourselves. The idea behind FEniCS is to
automate code generation so that the user can spend more time on doing
research and less time on implementation of assembly matrices. At the
core of FEniCS is DOLFIN \cite{logg2012dolfin}, which is C++/Python
library, and works as the main interface in FEniCS. In this thesis
only the Python interface has been used, in which C++ code is
automatically generated using SWIG. This allows for simplicity through
the Python scripting language and the speed of the C++ language.
The domain specific language used to represent weak formulations is
called the Unified form language (UFL) \cite{alnaes2014unified}, and
allows for e.g automatic differentiation of forms and expressions. The
FEniCS form compiler (FFC) \cite{logg2012ffc} compiles code written in
UFL to Unified Form-assembly Code (UFC) \cite{alnaes2012ufc} which are
optimized C++ code. The Python interface also makes use of the Instant
module which allows for just-in-time (JIT) compilation of C++
code. The compiled code is also stored in a cache so that compilation
of a form only happens once. Also, the relatively new UFL Analyser and
Compiler System (UFLACS) allows for fast compilation of complex forms
such as forms such as variational formulations that include the Holzapfel
Ogden material model \eqref{eq:holzapel_full}.

@@

\subsubsection{Numerical considerations}
The solution of non-linear problems such as the one described here is
typically solved using methods like Newton's method. The convergence of
such methods depends on the initial guess, and if the
initial guess is too far from the solution, the solver might diverge.
Moreover, if the initial guess is close to the solution the
convergence rate is in general quadratic.

>>>>

\begin{align*}
  \int_{\partial \omega} \mathbf{t} \mathrm{d}s
  = \int_{\partial \omega} \sigma \mathbf{n} \mathrm{d}s
  = \int_{\omega} \nabla \cdot \sigma \mathrm{d}v,
\end{align*}
and by collecting the terms from \eqref{eq:cons_lin_mom}. We arrive at
Cauchys momentum equation
\begin{align}
  \nabla \cdot \sigma + \mathbf{b} =  \rho \dot{\mathbf{v}}.
  \label{eq:chauch_momentum_eq}
\end{align}
The contribution from the body force ($\mathbf{b}$)  and inertial term
($\rho \dot{\mathbf{v}}$) are negligible compared to the stresses
\cite{hunter1996kd,tallarida1970left, moskowitz1981effects}, which is
why the force balance equations is typically only stated as
\begin{align}
  \nabla \cdot \sigma = \mathbf{0}.
  \label{eq:momentum_simple_current}
\end{align}
Note that we have formulated the balance law in the current
configuration. An equivalent statement can be formulated in terms of
the reference configuration
\begin{align}
  \nabla \cdot \FPK = \mathbf{0}, 
  \label{eq:momentum_simple_reference}
\end{align}
where $\FPK$ is the first Piola-Kirchhoff stress tensor.

@@

\subsubsection{General requirements for the strain-energy density
  function}
\label{sec:strain_energy_req}
Some general requirements must hold for the strain-energy function.
First of all, we require that the reference state is stress free and
that the stored energy increases monotonically with the deformation. 
Formally this can be stated simply as 
\begin{align*}
  \Psi(\I) = 0 \; && \text{and} &&\; \Psi(\F) \geq 0.
\end{align*}
Moreover, expanding or compressing a body to zero volume would
require an infinite amount of energy, i.e
\begin{align*}
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& 0 \\
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& \infty
\end{align*}
We say that the strain energy should be objective, meaning that the
stored energy in the material should be invariant with respect to
change of observer. Formally we must have: \emph{given any positive symmetric
rank-2 tensor $\C \in \mathrm{Sym}$:}
\begin{align}
  \Psi(\mathbf{C}) = \Psi(\mathbf{Q}\mathbf{C}\mathbf{Q}^T), \; \forall \mathbf{Q} \in \mathcal{G} \subseteq \mathrm{Orth}.
\end{align}
Here $\mathrm{Orth}$ is the group of all positive orthogonal matrices.
If $\mathcal{G} = \mathrm{Orth}$ we say that the material is
isotropic, and otherwise we say that the material is anisotropic.
This brings us to another important issue, which is related to the
choice of coordinate-system. Having to deal with different
coordinate-systems, and mapping quantities from one coordinate-system
to another can results in complicated computations. Therefore it would be beneficial if we
could work with quantities which do not depend on the choice of
coordinate-system. Such quantities are called invariants. 
If the material is isotropic, the representation theorem for
invariants states that $\Psi$ can be expressed in terms of the
principle invariants of $\mathbf{C}$, that is $\Psi = \Psi(I_1, I_2,
I_3)$. The principle invariants $I_i, i=1,2,3$ are the coefficients in
the characteristic polynomial of $\mathbf{C}$, and is given by 
\begin{align}
  I_1 = \tr \C,  && I_2 = \frac{1}{2}\left[ I_1^2 - \tr(\C^2)\right] && \text{and} && I_3 = \det \C.
\end{align}
In the case when the material constitutes a transversely isotropic
behavior, that is the material has a preferred direction $\mathbf{a}_0$,
which in the case of the myocardium could be the direction of fiber
muscle fibers, we have
\begin{align*}
  \mathcal{G} = \{ \mathbf{Q} \in \mathrm{Orth}: \mathbf{Q}(\mathbf{a}_0\otimes\mathbf{a}_0)\mathbf{Q}^T\},
\end{align*}
with $\otimes$ being the outer product. In this case the strain-energy
density function can still be expressed through invariants. However,
we need to include the so called quasi-invariants, which are defined
as stretches in the local microstructural coordinate-system. The
transversely isotropic invariants are given by
\begin{align*}
  I_{4\mathbf{a}_0 } = \mathbf{a}_0 \cdot (\C \mathbf{a}_0) && \text{and} && I_5 = \mathbf{a}_0 \cdot (\C^2 \mathbf{a}_0).
\end{align*}
The invariants do have a physical interpretation. For instance, $I_3$
is related to the volume ratio of material during deformation, while
$I_{4\mathbf{a}_0 } $ is related to the stretch along the direction
$\mathbf{a}_0 $. Indeed the \emph{stretch} ratio in the direction
$\mathbf{a}_0$ is given by $\lambda_{\mathbf{a}_0} = | \F \mathbf{a}_0
|$ and we see that $I_{4\mathbf{a}_0 }  =  \mathbf{a}_0 \cdot (\C
\mathbf{a}_0) = \F \mathbf{a}_0 \cdot (\F \mathbf{a}_0) =
\lambda_{\mathbf{a}_0}^2$. For more details about invariants see e.g
\cite{holzapfel2009constitutive,liu1982representations}.

The theory of global existence of unique solutions for elastic problems
was originally based convexity of the free energy function.
An energy function $\Psi: \mathrm{Lin}^+ \rightarrow
\mathbb{R}^+$ is strictly \emph{convex} if for each $\F \in
\mathrm{Lin}^+$ and $\mathbf{H} \neq \mathbf{0}$ with $\det (\F +
(1-\lambda)\mathbf{H}) > 0$, we have
\begin{align}
  \Psi(\lambda \F + (1-\lambda) \mathbf{H})
  < \lambda \Psi(\F)
  + (1-\lambda) \Psi(\F + \mathbf{H}), && \lambda \in (0,1).
\label{eq:strain_convex}
\end{align}
If the response $\FPK$ is differentiable, then condition
\eqref{eq:strain_convex} is equivalent of saying that the response is
positive definite, 
\begin{align}
  \mathbf{H} : \frac{\partial \FPK}{\partial \F} : \mathbf{H} > 0,
  && \F \in
     \mathrm{Lin}^+, \mathbf{H} \neq \mathbf{0}.
\label{eq:response_posdev}
\end{align}

However, from a physical point of view this requirement is too strict
\cite{ball1976convexity}. A slightly weaker requirement is the strong
ellipticity conditions which states that \eqref{eq:response_posdev}
should hold for any $\mathbf{H}$ of rank-one, which is analogous to
say that the strain energy function is rank-one convex.

% that the
% strain-energy function $\Psi$ is \emph{polyconvex}, meaning that there exist
% a convex function $\phi$ such that
% \begin{align*}
%   \Psi(\F) = \phi(\F, \cof \F, \det \F).
% \end{align*}
% Note however, that if $\Psi$ is convex then it is also polyconvex.

@@

Another approach which is used in this thesis is to impose a Robin
type boundary condition at the base
\begin{align}
  \FPK \Nvec + k \uvec = \mathbf{0},  \;  \Xvec \in \lvbase, 
\end{align}
or at the epicardium to mimic the pericardium
\begin{align}
  \FPK \Nvec + k \uvec = \mathbf{0},  \;  \Xvec \in \lvepi.
\end{align}
Here $k$ can be seen as the stiffness of a spring that limit the
movement. The limiting cases, $k = 0$ and $k \rightarrow
\infty$ represent free and fixed boundary respectively.
More complex boundary conditions to mimic the pericardium is also
possible \cite{fritz2014simulation}, but not considered in this thesis.
An overview of the location of the different boundaries for the
bi-ventricular geometry is illustrated in Figure \ref{fig:boundaries}.

@@

\subsection{Force-balance equation}
We will now collect all the terms that makes up to force balance for
the cardiac mechanics problem. Considering the myocardium as an incompressible,
hyperelastic material we obtain the following strong form in the
Lagrangian formulation 
\begin{align}
  \begin{split}
  \nabla \cdot \FPK &= 0 \\
  J - 1 &= 0,
  \end{split}
 \label{eq:force_balance_strong}
\end{align}
completed with appropriate boundary conditions. To solve this
numerically using the finite element method, we need to derive weak
variational form of this equation.
% \begin{align}
%   \FPK \Nvec + k \uvec &= \mathbf{0},  \;  \Xvec \in \lvbase\\
%   u_1 &= 0,  \;  \Xvec \in \lvbase \\
%   \FPK\mathbf{N}  &= \mathbf{0}, \;  \Xvec \in \epi \\
%   \FPK\mathbf{N} &= -p_{\mathrm{lv}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \lvendo
%   \label{eq:bndry_conditions_strong}
% \end{align}

@@

external energy. Internal energy is energy that is stored within the

@@

For an incompressible, hyperelastic material the total potential

@@

Equation \eqref{eq:force_balance_strong} is only possible to solve
analytically for very simplified problems. Therefore we need to employ
numerical methods to solve the problem. One such method
is the finite element method (FEM). When using the finite element method, we often refer to such
approximation as a Galerkin approximation. This is based on
approximating the solution by linear combinations of  basis functions in a finite dimensional
subspace of the true solution. If $V$ and $Q$ are two suitable Hilbert
spaces for the displacement $\uvec$ and the hydrostatic pressure $p$
respectively, we now select some finite dimensional subspaces $V_h
\subset V$ and $Q_h \subset Q$, which are spanned by a finite number
of basis functions.

@@

\begin{remark}
  The basis functions that span the Taylor-Hood
  finite element spaces are also known as the
  Lagrangian basis functions. These basis functions, of
  degree $n$, are polynomials of degree $n$ on each element, but only
  continuous at the nodes (i.e not continuously
  differentiable). Consequently, differentiating a function that is
  expressed as a linear combination of the Lagrangian basis functions,
  will not be continuous at the nodes, and therefore caution has to
  be made when evaluating features that depends on the derivative of
  such functions. Examples of such features are stress and
  strain with depends on the deformation gradient which again depends
  on the derivative of the displacement. One way to deal with this
  issue is to 1) use other types of elements that are continuously
  differentiable everywhere,  such a the cubic Hermite elements or 2) evaluate the features at the
  Gaussian quadrature points where there is no problem with continuity.
\end{remark}

\subsection{Constitutive relations}
\label{sec:constitutive_relations}
We have now covered a mechanical framework which holds any for
material in general. What differentiate the mechanics of soft
living tissue, like the myocardium, from other materials is the
constitutive relations which describes the response of a material to
applied load. Such constitutive relations often comes from
experimental observations, both observations of anatomical structure
but also from experiments done on tissue slabs.

@@

In this thesis we have in used a transversely isotropic version of
\eqref{eq:holzapel_full} which is obtained by setting $a_{fs} =
b_{fs}= a_s = b_s = 0$, i.e 
\begin{align}
  \label{eq:holzapel_trans}
  \begin{split}
  \Psi(I_1, I_{4\ef}) = \frac{a}{2 b} \left( e^{ b (I_1 - 3)}  -1 \right)
  + \frac{a_f}{2 b_f} \left( e^{ b_f (I_{4\ef} - 1)_+^2} -1 \right).
  \end{split}
\end{align}
If we further set $a_f = b_f = b = 0$ so that in $a$ is the
only nonzero parameter, then the Holzapfel-Ogden model reduces (after a
series expansion of the exponential and a limiting argument) to
\begin{align}
  \Psi(I_1)  = \frac{a}{2} \left( I_1 - 3 \right), 
\end{align}
which is the model of a Neo Hookean material. The Cauchy stress can be derived
analytically from \eqref{eq:holzapel_full}, by using the chain rule and
\eqref{eq:cauchy_incomp}, 
\begin{align}
  \begin{split}
    J\Cauchy
    =& \frac{\partial \Psi(\F)}{\partial \F}\F^{T} + p \I
    = \sum_{i \in \left\{ 1, 4\ef,  4\eS,  8\ef\eS \right\} }
    \psi_i \frac{\partial I_i}{\partial \F}\F^{T} + p \I \\
    =& p \I + a \left( e^{ b (I_1 - 3)}  -1 \right) \mathbf{B} 
    + 2 a_f (I_{4\ef} - 1)_+  e^{ b_f (I_{4\ef} - 1)^2} \mathbf{f} \otimes \mathbf{f} \\
    &+ 2 a_f (I_{4\eS} - 1)_+  e^{ b_f (I_{4\eS} - 1)^2} \mathbf{s} \otimes \mathbf{s} 
    + a_{fs} I_{8\ef\eS}  e^{ b_{fs} I_{8\ef\eS}^2} \left( \mathbf{f} \otimes \mathbf{s}  +  \mathbf{s} \otimes \mathbf{f} \right),
  \end{split}
\end{align}
where $\mathbf{B} = \F \F^{T}$ is the left Cauchy-Green tensor,
$\mathbf{f} = \F \ef$ and $\mathbf{s} = \F \eS$.

@@

The active stress approach is based one the classical three element
Hill model illustrated in Figure \ref{fig:hill_muscle_model}, where the
active contribution naturally decomposes the total stress into a sum
of passive and active stresses
\cite{nash2004electromechanical}. Hence, in the active stress
formulation \cite{hunter1998modelling} one assumes that the total
Cauchy stress $\Cauchy$ can be written as an additive sum of one
passive contribution $\Cauchy_p$ and one active contribution $\Cauchy_a$,
\begin{align}
  \Cauchy = \Cauchy_p + \Cauchy_a && J \Cauchy_p =  \frac{\partial \Psi(\F)}{\partial \F} \F^{T}.
\end{align}
The passive contribution is determined by the material model used,
while the active contribution is given by 
\begin{align}
  \Cauchy_a = \Cauchy_{ff} \Fef \otimes \Fef +
  \Cauchy_{ss} \mathbf{s} \otimes \mathbf{s} +
  \Cauchy_{nn} \mathbf{n} \otimes \mathbf{n},
\end{align}
and the different constants $\Cauchy_{ff}, \Cauchy_{ss}$, and
$\Cauchy_{nn}$, which are the active stress in the fiber, sheet and
sheet-normal direction respectively, are typically coupled to the
electrophysiology and calcium dynamics.
There are experimental evidence that the active stresses in the
transverse direction of the fibers ($\Cauchy_{ss}$, and $\Cauchy_{nn}$),
are non-negligible \cite{lin1998multiaxial}, and one approach is to assume
a uniform transverse activation in which the total active tension
can be written as 
\begin{align}
  \Cauchy_a = T_a \left[\Fef \otimes \Fef +
   \eta\left( \mathbf{s} \otimes \mathbf{s} +
  \ \mathbf{n} \otimes \mathbf{n} \right)\right],
  \label{eq:intro_active_stress}
\end{align}
where $\eta$ represent the amount of transverse activation and $T_a
\in \mathbb{R}$ is the magnitude of the active tension.
In the limiting case ($\eta = 0.0$), the active tension acts purely
along the fibers and \eqref{eq:intro_active_stress} reduces to 
\begin{align}
  \Cauchy_a = T_a \Fef \otimes \Fef.
\end{align}
Note that, by observing  that
\begin{align*}
  \frac{\partial I_{4\mathbf{a}_0}}{\partial \F}
  = \frac{\partial (\mathbf{a}_0  \cdot \C \mathbf{a}_0 )}{\partial \F}
  = 2 \mathbf{a} \otimes \mathbf{a}_0 \implies
  \mathbf{a} \otimes  \mathbf{a}= \frac{1}{2} \frac{\partial I_{4\mathbf{a}_0}}{\partial \F} \F^{T}
\end{align*}
and that $I_1 =  I_{4\ef} +  I_{4\eS} +  I_{4\en}$, 
we can instead decompose the strain-energy function into a passive and active
parts \cite{pathmanathan2010cardiac}, $\Psi= \Psi_p + \Psi_a$, with
\begin{align}
\Psi_a = \frac{T_a}{2J} \left(( I_{4\ef} - 1)  + \eta \left[ (I_1 - 3) -
    (I_{4\ef} - 1)\right] \right), 
\end{align}
so that $J \Cauchy_a  = \frac{\partial \Psi_a}{\partial \F}
\F^{T}$.

@@

The active part $\F_a$, is an inelastic process driven by the
biochemistry and can be seen as the actual distortion of the
microstructure. The elastic part $\F_e$ is responsible for preserving
compatibility of the tissue and stores all the energy in the
deformations. As a consequence, the strain energy function is a
function of the elastic deformation gradient only. The
decoupling can illustrated by considering two sarcomeres connected in
series as shown in Figure \ref{fig:actstrain}.

@@

The FEniCS project is an open-source computing platform for solving
partial differential equations (PDEs) using the finite element method (FEM).
Solving PDEs using FEM involves many implementation details that can
be tedious to implement yourselves. The idea behind FEniCS is to
automate code generation so that the user can spend more time on doing
research and less time on implementation of assembly matrices. At the
core of FEniCS is DOLFIN \cite{logg2012dolfin}, which is C++/Python
library, and works as the main interface in FEniCS. In this thesis
only the Python interface has been used, in which C++ code is
automatically generated using SWIG. This allows for simplicity through
the Python scripting language and the speed of the C++ language.
The domain specific language used to represent weak formulations is
called the Unified Form Language (UFL) \cite{alnaes2014unified}, and
allows for e.g automatic differentiation of forms and expressions. The
FEniCS form compiler (FFC) \cite{logg2012ffc} compiles code written in
UFL to Unified Form-assembly Code (UFC) \cite{alnaes2012ufc} which are
optimized C++ code. The Python interface also makes use of the Instant
module which allows for just-in-time (JIT) compilation of C++
code. The compiled code is also stored in a cache so that compilation
of a form only happens once. Also, the relatively new UFL Analyser and
Compiler System (UFLACS) allows for fast compilation of complex forms
such as forms such as variational formulations that include the Holzapfel
Ogden material model \eqref{eq:holzapel_full}.

@@

\subsubsection{Numerical considerations}
The solution of non-linear problems such as the one described here are
typically solved using methods like Newton's method. The convergence of
such methods depends on the initial guess, and if the
initial guess is too far from the true solution, the solver might diverge.
Moreover, if the initial guess is close to the true solution the
convergence rate is in general quadratic.
